---
title: 'Generalized Birthday Problem: Rcpp'
author: Chang-Yu Chang
date: '2018-03-30'
slug: generalized-birthday-problem-rcpp
categories:
  - R
tags:
  - academic
  - fun
  - R
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```
In the previous [post](https://changyuchang.netlify.com/post/generalized-birthday-problem/), I tried to use some algorithms to speed up simulation. Now in this post, I will introduce a pretty powerful package `Rcpp`, which has the potential to fundementally change the way you do simulation if you want to employ a time-consuming simulation in R. Let's finish this project!

- [Rcpp](https://changyuchang.netlify.com/post/generalized-birthday-problem-rcpp/#rcpp)

- [(Most) Generalized birthday problem ]()

### Rcpp
`Rcpp` is basically a package that incorporates C++ code in R. For me, using `Rcpp` means my code can work faster while I can still enjoy the convenience of many versatile R packages. A simple way to use `Rcpp` is the function `cppFunction`.  C++ code comes in as text in your R script. In this example, I define the function `simulateC`, which simulates and calculate the probility of a birthday shared by at least *k* persons in a room size of *room* (sorry for inconsistent notation, but I prefer i as index in a loop). I am totally not an expert in C++, so I am not goint through the code (this script is my first C++ script, to be honest). This is just to give a taste of how fast a function created by `Rcpp` can be.


```{r, eval=FALSE}
install.packages('Rcpp')
```
```{r}
library(Rcpp)
cppFunction('
#include <Rcpp.h>
using namespace Rcpp;
// [[Rcpp::export]]
float simulateC (int k, int room, int replicate) {
  IntegerVector result (replicate); 
  for (int r=0; r<replicate; r++){
    IntegerVector counts(365);
    for (int i=0; i<room; i++) {
      int pos = rand() % 365;
      if (pos < 365 && pos >= 0) counts[pos] ++;
    } 
    if (max(counts) >= k) {
      result[r] = TRUE;
    } else result[r] = FALSE;
  }
  float prob = float(sum(result))/float(replicate);
  return prob;
}
')
```

Let's benchmark the time spent by either `tabulate` or `Rcpp` method. `rbenchmark` is a package that we can benchmark the computer time. 

```{r}
library(rbenchmark)
k <- 2
room <- 23
r <- 1e4
b <- benchmark("Rcpp" = {simulateC(k=2, room=room, replicate=r)},
               'Tabulate' = {
                 birth <- matrix(sample(1:365, r*room, replace=TRUE), nrow=r, ncol=room)
                 sum(apply(birth, 1, function(x) max(tabulate(x))>=k)) /r},
               replications = 1000,
               columns=c('test', 'elapsed', 'replications', 'relative')
)

b 
```
Bascially, `benchmark` runs each methods for 1000 times, and compare the time consumed. It takes on average 11.072/1000 = 0.011072 seconds for running `simulateC(k=2, room=23, replicate=1e4)` once and 0.050681 seconds for running the `tabulate` method once. So on average, the `Rcpp` method is 4.5 times faster than `tabulate` metods. 

### Generalized Birthday Problem
This is the finall verison of function in generalized birthday problem project. 
```{r}
library(Rcpp)
gbp_rcpp <- function(k=2, p=.5, B_l=1000, B_u=20000, threshold=.001) {
  i <- k                       # Tested room size; start from k persons in a room
  x <- 1                       # Counter for while loop
  L <- 0                       # Lower bound for binary search
  R <- k^4+100+sample(-k:k, 1) # Upper bound for binary search
  
  # Create an empty data frame for result
  result <- setNames(data.frame(matrix(0, 100, 3)), c("room", "p", "replicate"))
  
  while (abs(result$room[x] - p) >= threshold) {
    # Increase replicate when approaching target room size
    r <- floor(-B_u/(1+exp(-.03* ((R-L)-200))))+(B_u+B_l)
    result$replicate[x] <- r
    
    # Test room size i
    i <- ceiling((L+R)/2)
    result$room[x] <- i 

    # Calculate probability by running simulation in C++
    result$p[x] <- simulateC(k=k, room=i, replicate = r)
    if (result$p[x] >= p) R <- i
    if (result$p[x] < p) L <- i
    
    # Break the loop if trapped
    if(sum(duplicated(tail(result$room[1:x], 2))) == 1 & result$room[x] != 0) break
    
    # Print out room sizes tested
    cat(paste0('Room size tested:', i, '\tprobability:', round(result$p[x], 3), '\treplicate size: ', r, '\n'))
    
    # Counter for steps
    x <- x + 1          
 }
  ans <- R
  return(list(answer=ans,
              result=result[1:x,]))
}

system.time(y_rcpp <- gbp_rcpp(k=8, p=.5, B_u=1000000))

```





